---
title: "Basic Statistics with R"
date: 
output: 
  html_document: 
    code_folding: hide
    highlight: zenburn
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
    
```
### Desriptive Statistics 
Organizing, summarizing and presentind the data, mean, median, mode..

## Inferntial Statistics
Drawing conclusions about a population based on data observed in a sample, confidence interval, test of significance

### Statistical Investigation
* Identify the problem
* Identify the objective of the study
* Identify the type of study
* Create a sampling plan
* Collect data
* Perform data analysis and interpretation
* Conclude the finding of the study


### Sampling Methods

| Sampling Method        | Description                                                                                                                                 |
|------------------------|---------------------------------------------------------------------------------------------------------------------------------------------|
| Simple Random Sampling | Each sample of the same size has an equal chane of being selected.                                                                          |
| Stratified Sample      | Divide the population into groups called starta and then take a sample from each stratum.                                                   |
| Cluster Sample         | Divide the population into strata and then randomly select some of the strata. All the members from these strata are in the cluster sample. |
| Systematic Sample      | Randomly select a starting point and take a very n-th piece of data from a list of the population                                           |

There are two types of random variables, **discrete** and **continous**.  
A discrete distribution has a finite number of values e.g. price of coke, age of employees.  
A continous distribution has all possible values in some range, sales per week, weight of students.

## Types of Distribution  
#### Discrete Probability Distribution  

* Binomial distribution
* Negative Binomial distribution
* Geometric distribution   

#### Continous Probability Distribution  

* Normal distribution
* Exponential distribution
* Gamma distribution
* Beta distribution

#### Normal distribution characterstics 
* Mean = Median = Mode  
* Symmetry about the center  
* 50% of values less than the mean and 50% greater than the mean    

Suppose 95% of the students at school are between 1.1m and 1.7m tall. Assuming this data is *normally distributed* can you
calculate the mean and the standard deviation?  
The mean is between halfway between 1.1 and 1.7:  
```{r}
meanofStudents <- (1.1 + 1.7)/2
cat("The mean:",meanofStudents)

```

95% is 2 standard deviation either side of the mean (a total of 4 standard deviations), so
```{r}
one_standard_deviation = (1.7 - 1.1)/4
cat("The SD:",one_standard_deviation)
```

* likely to be within 1 sd (68 out of 100 should be).
* very likely to be within 2 sd ( 95 out of 100 be).
* almost certainly with in 3 sd (997 out of 100 be).

### Z-Score

The number of standard deviations from the mean is also called the `z-score`.  
Example: In the above example in that same school one of your friends is 1.85m tall.  

* How far 1.85m from the mean? 
```{r}
far_from_mean <- 1.85 - 1.4 
far_from_mean
```

* How many standard deviations is that? The sd is 0.15:
```{r}
far_from_sd <-  0.45/0.15
far_from_sd 
```

Most students didn't even get 30 out of 60 and most will fail in the test. The test must have been really hard, so the professor dicedes to Standardize all the scores and only fail students 1 standard deviation below the mean.

```{r}
student_mark <- c(20,15,26,32,18,28,35,14,26,22,17)
mean_mark <- mean(student_mark)
sd_mark <- sd(student_mark)
cat("Mean of Score is:",mean_mark,"\nSD is:",sd_mark)

```

Now let's standardize the students mark:
```{r}
standardize_marks <-(student_mark - mean_mark)/6.6
standardize_marks

# 68% of the student mark will lay between 23-6.95 and 23+6.95
23 - 6.95 # lower bound
23 + 6.95 # upper bound

student_mark[student_mark < 16.05] 

```

A set of middle school students heights are normally distributed with `mean` of 150cm and a `sd` of 20cm. Let H be the height of a randomly selected student from the this set, find and interprate p(H > 170)?
```{r}

 1- 0.84 # 50% of the data and 34% from the half 1 standard deviation

# we can do by calculating
pnorm(170, mean = 150, sd = 20, lower.tail = F)
```

Suppose IQ's are normally distributed with  a mean of 100 and sd of 15. What percentage of people have IQ less than 125?
```{r}
pnorm(125, mean = 100, sd = 15, lower.tail = TRUE)
```

What is the percentage of people have an IQ greater than 110?
```{r}
pnorm(110, mean = 100, sd = 15, lower.tail = FALSE)
```

What percentage of people have an IQ between 110 abd 125?
```{r}
a <- pnorm(110, mean = 100, sd = 15, lower.tail = TRUE)
b <- pnorm(125, mean = 100, sd = 15, lower.tail = TRUE)

abs(a - b)

```

`qnorm` function is simply inverse of `pnorm`. You can use `qnorm` to determine the answer to the questions :  
What is the `z-score` of the pth quantile of the normal distribution?  

Suppose IQ's are normally distributed with a mean of 100 and standard deviation of 15.  
What IQ separates the lower 25% from the others?
```{r}
prob_25 <- qnorm(0.25, mean = 100, sd = 15, lower.tail = TRUE)
prob_25
```

What IQ separates the top 10% from the others?
```{r}
prob_10 <- qnorm(0.1, mean = 100, sd = 15, lower.tail = FALSE)
prob_10
```

The test score of an entrance exam fit a normal distribution with the mean test score of 72 and the sd of 15.2.  
Let's compute the percentage of students scoring 84 or more?
```{r}
pnorm(84, mean = 72, sd = 15.2, lower.tail = FALSE)
```

### Skewness 
* Skewnes > 0: 
  Right skewed distribution-most values are concentrated on the left of the mean, with extrem values to the right.

* Skewness < 0 :  
  Left skewed distribution-most values are concentrated on the right of the mean, with extreme values to the left.

* Skewness = 0: mean = median: 
  The distribution is symetrical aroun the mean.

### Kurstosis 
  - Kurtosis indicates how the peak and tails of a distribution differ from the normal distribution.  
  - Use kurtosis to help you initially understand general characterstics about the distribution of your data.
  - *positive kurtosis*: A value indicates that the distribution has heavier tails and a shaper peak than the normal distribution.
  - *negative kurtosis*: A value indicates that the distribution has ligther tails and a flatter peak than the normal disttibution. 
  
## Practical Application - Central Limit Theorm
The mean salary of the 9,000 employees at Holley.com is $\mu$ = 26,000 with standard deviation of $\sigma$ = 2,420. A pollster samples 400 randomly selected employees and finds that the mean salary of the sample is `26,650`. Is it likely that the pollster would get these results by chance, or does the discripancy suggest that the pollester's results are fake?

_Solution_

  - population Mean = 26,000
  - population sd = 2,420
  - As per central limit theorm,
    - sample size = 400, mean of sample distribution of mean, $\mu$ = 26,000
    - sample standard deviation = $\sigma$ / $\sqrt{n}$  
z = (data value - mean) / sd
```{r}
z = (26650 - 26000)/121
z
```
Mean salary is more than 5 standard deviations above the mean salary of the entire company.  
According to the z-score, a z-score of 5.37 lies beyond the 99.99th percentile.  

That is, the likelihood of selecting a group of 400 workers with a mean salary 26,650 is about less than 0.01%.  
```{r}
pnorm(26650, mean = 26000, sd = 121, lower.tail = FALSE)
```

The mean salary of the sample is surprisingly high, perhaps the survey was flawed.  

Engineers must consider the breadths of male heads when designing motorcycle helmets for men. Men have have head breadths that are normally distributed with mean 6.0 inches and standard deviation of 1.0 inch.  

  - If 1 man is randomly selected from the population, find the probability that his head breadth is less than 6.2 inch.
  - 57.93% have head size less than 6.2 inches.
  
```{r}
pnorm(6.2, mean = 6, sd = 1, lower.tail = TRUE)

```
  - The safeguard helmet company plans an initial production run of 100 helmets. How likely is it that 100 randomly selected men have a mean head breath of less that 6.2 inches?
```{r}
sample_sd = 1/sqrt(100)
pnorm(6.2, mean = 6, sd = 0.1, lower.tail = TRUE)
```
  
  - 98% have head size less than 6.2 inches 
  
ACME motorcycle company is making a new adjustable helmet. In reality, it is not yet economical to make a helmet that fits everyone. You must desing a helmet that will fit all but largest 5% fo male head breadths.  
What is the largest size male head breadth that your new helmet will fit?  
```{r}
qnorm(0.05, sd= 1, mean = 6, lower.tail = FALSE)
```
  - The largest male head breadth will be 7.64 inches.   

An icecream campany claims that maximum saturated fat content in chip packet is 2 gram with sd of 0.25. At test on a sample of 35 packets is done. How likely is that the mean saturated fat of the 35 packets would be more than 2.1 gram?
```{r}
sample_sd = 0.25 / sqrt(35)
pnorm(2.1, mean = 2, sd = sample_sd, lower.tail = FALSE)

```
